Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-09-09 README.Rmd
small updates after meeting

1. Pull the repository from GitHub.
2. Unzip all .zip files in the repository (`data-bullet-train.csv.zip`, `data-bullet-test.csv.zip`, `data-example-signatures.csv.zip`, and `figure-08-1.png.zip`).
3. Install R packages goodekat/lime and goodekat/limeaid from GitHub using devtools. (Note that goodekat/lime is a forked versions of thomasp85/lime with minor changes to extract objects for the analysis in the manuscript.)

```{r eval = FALSE}
devtools::install_github("goodekat/lime")
devtools::install_github("goodekat/limeaid")
```

4. Knit paper (via the paper.Rnw file).

## Additional Information on Bullet Data

>>>>

1. Pull the repository from GitHub.
2. Install R packages goodekat/lime and goodekat/limeaid from GitHub using the remotes R package. (Note that goodekat/lime is a forked versions of thomasp85/lime with minor changes to extract objects for the analysis in the manuscript.)

```{r eval = FALSE}
remotes::install_github("goodekat/lime")
remotes::install_github("goodekat/limeaid")
```

3. Knit paper (via the paper.Rnw file).
et
## Additional Information on Bullet Data

### 2020-09-09 paper.Rnw
small updates after meeting

% Abstract and keywords
\abstract[Abstract]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations). LIME uses a surrogate model to explain the relationship between predictor variables and predictions from a black-box model in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce three visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplots, (2) assessment metric plots, and (3) feature heatmaps. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameter values and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made of LIME that are similar to other criticisms in the literature.}
\keywords{LIME, black-box models, interpretability, diagnostics}

@@

# Load packages
library(Cairo)
library(caret)
library(cowplot)
library(dplyr)
library(forcats)
library(ggforce)
library(ggplot2)
library(ggpcp) #devtools::install_github("yaweige/ggpcp")
library(glmnet)
library(gower)
library(graphics)
library(gretchenalbrecht) #devtools::install_github("dicook/gretchenalbrecht")
library(latex2exp)
library(lime) #devtools::install_github("goodekat/lime")
library(limeaid) #devtools::install_github("goodekat/limeaid")
library(magick)
library(purrr)
library(randomForest)
library(stringr)
library(tidyr)

@@

\item \emph{Explainer Model Fitting}: Apply the black-box model to the simulated data to obtain predictions. Compute the distance between each of the simulated observations and the prediction of interest. Perform feature selection. Fit an interpretable model with the black-box predictions from the simulated data as the response, the selected features from the transformed simulated data as the predictors, and the distances as weights. This model is the explainer model.

@@

# Load the bullet matching training data and make some adjustments:
#   1. filter out observations known to have tank rash
#   2. exclude features not used to train the rtrees random forest 
#   3. add a case variable
#   4. change the name of the same_source variable to match the test data
#   5. reorder the variables
bullet_train <-
  read.csv("data-bullet-train.csv") %>%
  filter(!(
    land_id1 %in% c(
      "Hamby173-Br6-B2-L1",
      "Hamby173-Br9-B2-L4",
      "Hamby173-BrUnk-BB-L2",
      "Hamby173-BrUnk-BQ-L4"
    ) |
      land_id2 %in% c(
        "Hamby173-Br6-B2-L1",
        "Hamby173-Br9-B2-L4",
        "Hamby173-BrUnk-BB-L2",
        "Hamby173-BrUnk-BQ-L4"
      )
  )) %>%
  select(-lag, -abs_lag, -signature_length) %>%
  mutate(case = factor(1:n())) %>%
  rename("samesource" = "same_source") %>%
  select(case, land_id1, land_id2, everything())

# Load the bullet test data
bullet_test <- read.csv("data-bullet-test.csv")

@@

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). \kge{See the supporting information for additional information on the data.} The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes \Sexpr{dim(bullet_train)[1]} land-to-land comparisons.

@@

# Convert the bullet figure to eps (if needed)
if (!file.exists("figure-08-1.eps")) {
  bullets <- image_read("figure-08-1.png")
  image_write(bullets, path = "figure-08-1.eps", format = "eps")
}

@@

# Load the signatures data
signatures <- read.csv("data-example-signatures.csv")

@@

\kge{The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}. Additional information about the bullet matching data are available at \kgc{XXX}.}

@@

For $x_e$ and a set of tuning parameter values $t$, let $\textbf{X}_{e,t}'$ be the LIME simulated dataset with $K$ features and $S$ observations such that $x'_{e,t,s}$ is the feature vector for simulated data observation $s$ corresponding to observation/explanation $e$ and tuning parameter values $t$. Let $\textbf{Z}'_{e,t}$ be the matrix of simulated data transformed to bin indicator variables (for bin based simulation methods) or standardization (for density based simulation methods)  with $K$ features and $S$ observations such that $z'_{e,t,s}$ is the interpretability transformed feature vector for observation/explanation $e$, tuning parameter values $t$, and simulated data observation $s$. Note that $z_{e,t}$ will represent the  transformed version of $x_e$.

Next, let $\omega_t$ represent a proximity distance metric corresponding to tuning parameter values $t$. Then $\omega_t\left(x_e, x'_{e,t,s}\right)$ is the weight assigned to $x'_{e,t,s}$, which is the proximity between $x_e$ and $x'_{e,t,s}$. Allow $g_{e,t}$ to be the explainer model for an explanation $e$ and tuning parameter values $t$. Thus, $g_{e,t}\left(z'_{e,s,t}\right)$ is the explainer model prediction for the interpretability transformed simulated data observation $s$.

@@

\noindent where $\mathcal{L}$ was originally defined in \citet{ribeiro:2016}.\\
\\
\emph{Mean squared explanation error} is denoted by MSEE and computed as

>>>>

% Abstract and keywords
\abstract[Abstract]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations). LIME uses a surrogate model to explain the relationship between predictor variables and predictions from a black-box model in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce three visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplots, (2) assessment metric plots, and (3) feature heatmaps. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameter values and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made of LIME that are similar to other criticisms in the literature.}
\keywords{explainable machine learning, black-box models, interpretability, statistical graphics, data science}

@@

# Load packages
library(Cairo)
library(caret)
library(cowplot)
library(dplyr)
library(forcats)
library(ggforce)
library(ggplot2)
library(ggpcp) #remotes::install_github("yaweige/ggpcp")
library(glmnet)
library(gower)
library(graphics)
library(gretchenalbrecht) #remotes::install_github("dicook/gretchenalbrecht")
library(latex2exp)
library(lime) #remotes::install_github("goodekat/lime")
library(limeaid) #remotes::install_github("goodekat/limeaid")
library(magick)
library(purrr)
library(randomForest)
library(stringr)
library(tidyr)

@@

\item \emph{Explainer Model Fitting}: Apply the black-box model to the simulated data to obtain predictions. Compute the distance between each of the simulated data points and the prediction of interest. Perform feature selection. Fit an interpretable model with the black-box predictions from the simulated data as the response, the selected features from the transformed simulated data as the predictors, and the distances as weights. This model is the explainer model.

@@

# Load the bullet matching training data and make some adjustments:
#   1. filter out observations known to have tank rash
#   2. exclude features not used to train the rtrees random forest 
#   3. add a case variable
#   4. change the name of the same_source variable to match the test data
#   5. reorder the variables
bullet_train <-
  readr::read_csv("data-bullet-train.csv.zip") %>%
  filter(!(
    land_id1 %in% c(
      "Hamby173-Br6-B2-L1",
      "Hamby173-Br9-B2-L4",
      "Hamby173-BrUnk-BB-L2",
      "Hamby173-BrUnk-BQ-L4"
    ) |
      land_id2 %in% c(
        "Hamby173-Br6-B2-L1",
        "Hamby173-Br9-B2-L4",
        "Hamby173-BrUnk-BB-L2",
        "Hamby173-BrUnk-BQ-L4"
      )
  )) %>%
  select(-lag, -abs_lag, -overlap, -signature_length) %>%
  mutate(case = factor(1:n())) %>%
  rename("samesource" = "same_source") %>%
  select(case, land_id1, land_id2, everything())

# Load the bullet test data
bullet_test <- readr::read_csv("data-bullet-test.csv.zip")

@@

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). See the supporting information for additional information on the data. The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes over 80,000 land-to-land comparisons.

@@

# Convert the bullet figure to eps (if needed)
if (!file.exists("figure-08-1.eps")) {
  unzip("figure-08-1.png.zip")
  bullets <- image_read("figure-08-1.png")
  image_write(bullets, path = "figure-08-1.eps", format = "eps")
}

@@

# Load the signatures data
signatures <- readr::read_csv("data-example-signatures.csv.zip")

@@

The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}. Additional information about the bullet matching data are available at XXX.

@@

For $x_e$ and a set of tuning parameter values $t$, let $\textbf{X}_{e,t}'$ be the LIME simulated dataset with $K$ features and $S$ rows such that $x'_{e,t,s}$ is the feature vector for simulated data point $s$ corresponding to explanation $e$ and tuning parameter values $t$. Let $\textbf{Z}'_{e,t}$ be the matrix of simulated data transformed to bin indicator variables (for bin based simulation methods) or standardization (for density based simulation methods) with $K$ features and $S$ observations such that $z'_{e,t,s}$ is the interpretability transformed feature vector for explanation $e$, tuning parameter values $t$, and simulated data point $s$. Note that $z_{e,t}$ will represent the  transformed version of $x_e$.

Next, let $\omega_t$ represent a proximity distance metric corresponding to tuning parameter values $t$. Then $\omega_t\left(x_e, x'_{e,t,s}\right)$ is the weight assigned to $x'_{e,t,s}$, which is the proximity between $x_e$ and $x'_{e,t,s}$. Allow $g_{e,t}$ to be the explainer model for an explanation $e$ and tuning parameter values $t$. Thus, $g_{e,t}\left(z'_{e,s,t}\right)$ is the explainer model prediction for the interpretability transformed simulated data point $s$.

@@

\noindent where $\mathcal{L}$ is the fidelity metric originally defined in \citet{ribeiro:2016}.\\
\\
\emph{Mean squared explanation error} is denoted by MSEE and computed as

### 2020-09-09 paper.tex
knit paper after changing file names
small updates after meeting

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). \kge{See the supporting information for additional information on the data.} The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes 83436 land-to-land comparisons.

>>>>

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). See the supporting information for additional information on the data. The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes over 80,000 land-to-land comparisons.
